Comparing Apples and Oranges in Youth Depression Treatments?

### 2024-10-04 manuscript_with_multilevel_metan.qmd
more cleaning and updated main qmd to match revision

---
title: "Comparing Apples and Oranges in Youth Depression Treatments?"
subtitle: "A Quantitative Critique of the Evidence Base and Guidelines"
author:
  - name: Argyris Stringaris
    email: a.stringaris@ucl.ac.uk
    affiliations: 
        - id: ucl_1
          name: University College London
          department: Divisions of Psychiatry and Psychology and Language Science
          address: 1-19 Torrington Place
          city: London, UK
          state: 
          postal-code: WC1E 7HB
          country: United Kingdom
    attributes:
        corresponding: true
    note: 
  - name: Charlotte Burman
    email: charlotte.burman@ucl.ac.uk
    affiliations:
        - ref: ucl_1
    note:
  - name: Raphaelle Delpech
    email: raphaelle.delpech.18@ucl.ac.uk
    affiliations:
        - ref: ucl_1
  - name: Rudolf Uher
    email: uher@Dal.Ca
    affiliations:
        - id: dal
          name: Dalhousie University
          department: Department of Psychiatry
          address: 
          city: Nova Scotia
          state: 
          postal-code: 
          country: Canada
  - name: Dayna Bhudia
    email: dayna.bhudia@ucl.ac.uk
    affiliations:
        - ref: ucl_1
  - name: Despoina Miliou
    email: dgmiliou@outlook.com.gr
    affiliations:
        - ref: ucl_1
  - name: Ioannis-Marios Rokas
    email: rokasgiannis93@gmail.com
    affiliations:
        - ref: ucl_1
  - name: Marinos Kyriakopoulos
  - name: Lucy Foulkes
    email: lucy.foulkes@psych.ox.ac.uk
    affiliations:
        - id: OU
          name: University of Oxford
          department: Experimental Psychology
          address: Anna Watts Building, Radcliffe Observatory Quarter, Woodstock Road
          city: Oxford
          postal-code: OX2 6GG
          country: United Kingdom
    note:
  - name: Carmen Moreno
    email: cmoreno@hggm.es
    affiliations:
        - id: HGM
          name: Hospital Gregorio Marañón
          department: Department of Psychiatry
          address: 46 C. del Dr. Esquerdo
          city: Madrid
          state:
          postal-code: 28007
          country: Spain
  - name: Samuele Cortese
    email: samuele.cortese@soton.ac.uk
    affiliations:
        - id: soton
          name: University of Southampton
          department: Centre for Innovation in Mental Health (CIMH), School of Psychology
          address: Highfield Campus, Building 44
          city: Southampton
          state:
          postal-code: SO17 1BJ
          country: United Kingdom
  - name: Glyn Lewis
    email: glyn.lewis@ucl.ac.uk
    affiliations:
        - ref: ucl_1        
  - name: Georgina Krebs
    email: g.krebs@ucl.ac.uk
    affiliations:
        - ref: ucl_1
    note:
keywords: 
  - depression
  - adolescents
  - antidepressants
  - psychotherapy
date: last-modified
format:
  docx:
    reference-doc: "bibliography/word-style-reference-doc.docx"
    fig-width: 9
    fig-height: 7
    fig_caption: yes
    tbl_caption: yes
    number-sections: true
execute: 
  echo: true
  auto: true
  warning: FALSE
bibliography: "bibliography/references.bib"
csl: "bibliography/vancouver_style.csl"
font:
  family: Calibri
fig-cap-location: top
---

@@

###########Libraries needed
library(tidyverse)
library(truncnorm)
library(metafor)
library(meta)
library(kableExtra)
library(flextable)
library(magrittr)
library(officer)
library(RefManageR)
library(ftExtra)
library(dplyr)
library(broom)
library(effsize)
library(here)

@@

# For referencing
bib_data <- ReadBib(here("multilevel_metaregression", "bibliography", "references.bib"))

@@

**Objectives:** Should a young person receive psychotherapy or medication for their depression, and on what evidence do we base this decision? In this paper, we test whether the basic conditions required to draw valid inferences to answer this question are currently met. **Methods:** We included 88 RCTs of psychotherapy and medication for child and adolescent depression (mean age 4-18 years). Using meta-analyses, we compared a) participant characteristics and b) trial characteristics in medication and psychotherapy trials. Lastly, we examined whether psychotherapy controls are well-matched to active conditions. **Results:** Participants in medication RCTs had higher depression severity and were more frequently male compared to psychotherapy RCTs. There was a dramatic difference in the within-subject improvement due to placebo (SMD=-1.9 (95% CI: -2.10 to -1.70)) vs psychotherapy controls (SMD=-0.5 (95% CI: -0.75 to -0.25)). Within psychotherapy RCTs, control conditions were less intensive on average than active conditions. **Conclusions:** Medication and psychotherapy RCTs differ on fundamental participant and methodological characteristics, thereby violating key conditions for valid comparison between them. Psychotherapy controls often involve little therapist contact and are easy-to-beat comparators. These findings cast doubt on the confidence with which psychotherapy is recommended for youth depression, and highlight the pressing need to improve the evidence base.

@@

Psychotherapy is recommended before medication for most cases of depression in children and adolescents, a recommendation that is based on indirect comparisons of outcomes from randomised controlled trials (RCTs) within each treatment modality. We examine the validity of these inferences by scrutinising the comparability of psychotherapy and medication RCTs. We find significant differences in sample characteristics (namely depression severity and sex composition) and trial design features, such that the within-group effect sizes of medication controls (i.e. pill placebo) are much larger than those for psychotherapy controls, and that medication RCTs feature significantly more trial sites. We also examine the quality of controls used in psychotherapy RCTs and find that they are poorly matched to active intervention arms in ways such as human contact hours, and hence represent poor and easy to beat comparators. Our findings underscore the need for a higher quality evidence base upon which to base treatment guidelines and clinical decision making.

@@

For adolescent depression, there are limited head-to-head trials of medication and psychotherapy, and hence recommendations are derived from indirect comparisons of treatment efficacy. The National Institute of Health and Care Excellence (NICE) guidelines for adolescent depression recommend psychotherapy (specifically cognitive behaviour therapy (CBT) or interpersonal therapy (IPT)) over medication in most cases [@niceDepressionChildrenYoung2019]. This is in keeping with two sources of evidence relating to child and adolescent depression: meta-analyses of medication randomised controlled trials (RCTs) that cast doubt on the efficacy of antidepressants, with the exception of fluoxetine [@ciprianiComparativeEfficacyTolerability2016]; and meta-analyses of psychotherapy RCTs that conclude psychotherapy to be efficacious [@cuijpersEffectsPsychologicalTreatments2021]. However, a recent network meta-analysis (NMA) [@zhouComparativeEfficacyAcceptability2020], an established method of comparing treatments using both direct and indirect (i.e. treatment A with treatment C, via studies that directly compare A with B and B with C) evidence, concluded that only fluoxetine alone and fluoxetine administered together with CBT were significantly more effective than medication control (i.e. pill placebo) or psychotherapy controls. The only head-to-head RCT comparing modalities found that fluoxetine, alone and in combination with CBT, was superior to pill placebo, though CBT alone was not [@marchFluoxetineCognitiveBehavioralTherapy2004]. Also, the addition of psychotherapy to standard care did not improve outcomes [@goodyer_randomised_2008]. Given this confusing evidence base, how should patients, carers, clinicians and policy makers make treatment decisions?

In this paper, we examine whether the existing evidence for adolescent depression treatments can offer valid answers to this question. We provide a conceptual framework and test a series of hypotheses using data from existing trials. Two points are crucial to indirect comparisons of treatment modalities. First, whether the participants in trials are comparable across modalities. Second, whether key conditions of the trial, such as the effects of control conditions or the number of sites involved, are comparable.

Starting with the first point, comparison between different trials assumes that they sample from the same population. If not, the validity of any comparisons, including those conducted through NMA (which rests on the principal of transitivity, i.e. the requirement that the different sets of randomized trials are similar on average [@chaimaniChapter11Undertaking2023]) are questionable.

The assumption that medication and psychotherapy trials sample from the same population may not be valid as patients and parents often have treatment preferences [@jaycoxAdolescentPrimaryCare2006; @langerParentYouthPreferences2021; @mchughPatientPreferencePsychological2013], meaning that there is likely to be a self-selection bias in who participates in psychotherapy and medication trials. Moreover, treatment preferences correlate with clinically-relevant participant characteristics, including severity and sex. Some of these characteristics, such as severity, may moderate treatment response [@courtneyPredictorsModeratorsMediators2022; @lorenzo-luacesDoubleTroubleSymptom2020] and if they differ across psychotherapy and medication trials they may confound comparisons.

Regarding the second point, differences in trial design may impact outcomes in a differential way between medication and psychotherapy trials [@delgiovaneCombiningPharmacologicalNonpharmacological2019]. Most obviously, participants in psychotherapy trials are generally unblinded to treatment allocation, with the exception perhaps of trials that compare two equally plausible treatment arms [@calvoInterventionAdolescentsEarlyOnset2014]. This creates differential expectations which may favour the psychotherapy active condition, as participants are content to be receiving the "cutting edge" treatment, whilst those in the control are dissatisfied for having missed out (i.e. "disappointment bias" [@reltonEthicsTrialsCohorts2017]. By contrast, in new antidepressant trials, patients (and raters) were largely unable to judge treatment allocation [@linAssessmentBlindingRandomized2022], suggesting that expectancy effects are well-matched across conditions. Since expectancy is substantially associated with treatment outcomes [@constantinoExpectations2011], if expectancy differs between medication and psychotherapy trials, comparisons between them, including in NMA, become questionable.

Another difference in design is the number of trial sites. The number of sites in medication trials is positively related to the magnitude of placebo response [@bridgePlaceboResponseRandomized2009; @dechartresSingleCenterTrialsShow2011; @meisterPlaceboResponseRates2020]. This phenomenon may be due to lower quality of assessments in multi-site trials, with higher rates of classification errors and therefore higher apparent spontaneous remission or regression to the mean.

@@

Additionally, control conditions in RCTs should generate counterfactual conditions to the intervention [@guoChapterCounterfactualFramework2014]: what would have been the outcome had an individual not received the intervention, with all else being equal. Pill placebo, where the appearance of the drug is faithfully emulated, is an effort for all else to be equal. In psychotherapy trials, control conditions may not be so well matched to the intervention (e.g. in number of hours of therapist contact).

@@

In this study we examine RCTs of psychotherapy and medication for child and adolescent depression (mean age 4-18 years). We hypothesise that there are substantial differences between psychotherapy and medication RCTs, making their comparison problematic, which we examine in the following ways. First, we conduct meta-analyses to compare sample characteristics of medication and psychotherapy trials including: a) baseline depression severity; b) percentage females; and c) mean age. Second, we examine trial characteristics including the efficacy of the control arms, using random-effects meta-regression, and the number of trial sites. Third, we examine the quality of psychotherapy control conditions by scrutinising the extent to which they are matched to the active intervention in ways such as number and frequency of sessions, and hence whether they represent fair pairings from which to draw valid efficacy inferences.

@@

The study protocol was registered on the Open Science Framework (OSF) and can be found at <https://osf.io/bfmc6>.

@@

We included RCTs identified in a recent meta-analysis of psychotherapy versus control [@cuijpersEffectsPsychologicalTreatments2021] and an NMA examining the efficacy of antidepressants [@ciprianiComparativeEfficacyTolerability2016] for depression in children and adolescents. For the psychotherapy trials, we utilised open data from the previous meta-analysis (see <https://docs.metapsy.org/databases/depression-childadol-psyctr/>). For medication trials, we were unable to access the full dataset used in the NMA and hence extracted data from the included studies ourselves.

For medication trials, we also conducted a systematic search for studies published after the final search date of Cipriani et al.'s [@ciprianiComparativeEfficacyTolerability2016] review up to the final search date of Cuijpers et al's [@cuijpersEffectsPsychologicalTreatments2021] review to ensure we analysed an equivalently up-to-date database of medication trials. Two authors screened 450 titles and abstracts, and 38 full text records. Seven studies met inclusion criteria and one author completed data extraction for these papers (see @fig-prisma for PRISMA diagram).

@@

We conducted random-effects meta-analyses and tested subgroup differences (psychotherapy vs medication trials) in severity of depressive symptoms, sex and age. Meta-analyses were implemented using R's meta package.

@@

As the measure of effect of each individual study, we used the within-group Standardised Mean Difference (SMD) for the primary depression scale used (selected according to the hierarchy defined in the Supplement).

Where individual studies did not report all data required to calculate the SMD, we imputed missing data according to the methods summarised in the Cochrane Handbook [@higginsChapterChoosingEffect2023].

For meta-analysis it is necessary to estimate a standard error of the SMD. This requires a correlation between the pre- and post-measures, a statistic typically not reported. To ensure that our results are not biased by misestimation [@cuijpers_pre-post_2017], we simulated n=1000 datasets for different values (0.45 to 0.9) of this correlation and used these datasets in subsequent analyses (please refer to Supplement for full details).

@@

We estimated pooled SMDs for each arm by using multilevel models implemented in R's metafor package.

@@

We conducted a series of sensitivity analyses where we excluded studies that 1) used waitlist as their control and 2) recruited participants with subclinical levels of depression. Next, we conducted two analyses where we included only trials that used the Children's Depression Rating Scale, Revised (CDRS-R) or the Hamilton Depression Rating Scale (HAM-D) as outcome instruments. Additonally, we restricted two analyses to studies with variance below 0.02 and which reported SD at outcome respectively.

Further, we tested whether the simulated values for the standard error had a substantial influence on the estimation of the differences between the medication and psychotherapy control conditions. We plotted the z-value of the difference between the two coefficients against the number of simulations. We make inference on the stability of the difference, by counting the proportion of times that the z-value is above the critical value of z = `r x` corresponding to an alpha = 0.05.

@@

Data for included studies are summarised in @tbl-all-trials and are also available as a csv dataframe on \[<https://github.com/transatlantic-comppsych/apples_oranges>\]. Please see @fig-prisma for a summary of the sources of included RCTs.

In total, there were `r length(unique(df_long_for_metan$new_study_id))` RCTs which included `r sum(df_count_studies[df_count_studies$condition == "medication_active", "n"])` active arms and `r sum(df_count_studies[df_count_studies$condition == "medication_control", "n"])` control arms of medication trials; and `r sum(df_count_studies[df_count_studies$condition == "psychotherapy_active", "n"])` active arms and `r sum(df_count_studies[df_count_studies$condition == "psychotherapy_control", "n"])` control arms from psychotherapy RCTs. Note that the number of active and control arms does not exactly match because some studies feature more than one control or active arm.

Placebo pill was the control condition for all medication trials. In psychotherapy trials, the control arm included 14 waitlist, 25 treatment-as-usual (TAU), and 19 other control conditions (e.g. relaxation training, attention control, counselling).

@@

@tbl-baseline-results summarises the results from each of the meta-analyses examining sample characteristics at baseline. The summary statistics are provided for each subgroup (i.e. medication and psychotherapy) and the p-value derives from the test for subgroup differences.

@@

We conducted a series of sensitivity analyses (all figures in the Supplement). Excluding waitlist control studies (see @fig-plot-means-no-wl) and sub-clinical studies (see @fig-plot-means-clin) yielded a pattern of results very similar to the overall analyses. Next, we examined the data including only those studies that used the CDRS (see @fig-plot-means-cdrs) or the HAMD (see @fig-plot-means-hamd). Medication control and psychotherapy control conditions remained significantly different, though the small number of studies resulted in less precise estimates of the SMDs. Conversely, restricting the analysis to studies with variance below 0.02 yielded a similar pattern of results, but with increased precision in SMD estimates across conditions (see @fig-plot-means-var2). Finally, we showed that different values for the pre-post measure correlation had minimal effect on the estimated outcomes (see @fig-stab-sims).

@@

We addressed potential regression to the mean by including the baseline score for each depression scale in the linear regression model as per equation 3 in @barnettRegressionMeanWhat2004 (see @tbl-de-meaned in the Supplement). The difference between the medication control and psychotherapy control arms remained significantly different.

@@

Average number of trial sites was significantly higher in medication trials (*M* = `r df_sites_summary_stats$Mean[df_sites_summary_stats$Modality == "Medication"]`, *SD* =`r df_sites_summary_stats$SD[df_sites_summary_stats$Modality == "Medication"]`) compared to psychotherapy studies (*M* =`r df_sites_summary_stats$Mean[df_sites_summary_stats$Modality == "Psychotherapy"]`, *SD* =`r df_sites_summary_stats$SD[df_sites_summary_stats$Modality == "Psychotherapy"]`)(*t* (`r round (sites_t_test$parameter, 2)`) = `r round(sites_t_test$statistic,2)`, *p* =`r sites_t_test_stats$p_value`). Of those studies with data available, 26 of 28 (93%) medication trials were multisite, compared to 25 of 46 (54%) psychotherapy studies.

@@

As seen in @tbl-intensity-of-int, active conditions featured significantly more sessions when compared to control conditions. Sessions in active conditions were longer and more frequent, resulting in significantly more intervention hours overall. Notably, many control conditions were very poorly described and their intensity could not be quantified, resulting in missing data. We performed a sensitivity analysis where we excluded trials using waitlist controls; with the exception of number of sessions, differences between active and control arms were no longer statistically significant though remained substantial, with between-group Cohen's ds ranging from 0.41-0.54 (please see @tbl-intensity-of-int-no-wl in the Supplement).

@@

This paper sought to address the question of whether psychotherapy and medication can be meaningfully compared on the basis of the existing evidence, by looking at two key conditions for comparability. First, whether the participants of trials in one modality are comparable to those in another modality. Second, whether conditions of the trial, such as the effects of control conditions or the number of sites involved, are comparable.

Starting with the first question, we found that participants in medication trials are comparable on age but are more likely to be male and have more severe depression compared to those in psychotherapy trials. This indicates that different people enter medication and psychotherapy trials, thereby violating basic assumptions of comparability.

Severity is particularly important as it may moderate treatment response, with some evidence suggesting that those with higher baseline scores respond more to antidepressants [@stoneResponseAcuteMonotherapy2022] or that their response to pill placebo is lower [@bridgePlaceboResponseRandomized2009]. Other studies argue against severity as a treatment moderator [@trogerBaselineDepressionSeverity2024; @weitzBaselineDepressionSeverity2015], however these are within people who have chosen to be in the particular trial and modality. Moreover, severity may represent different subtypes in terms of course of depression and real-life outcomes [@lamersSixyearLongitudinalCourse2016; @simmonds-buckleyDepressionSubtypesTheir2021].

>>>>

---
title: "Comparing Apples and Oranges in Youth Depression Treatments?"
subtitle: "A Quantitative Critique of the Evidence Base and Guidelines"
author:
  - name: Argyris Stringaris
    email: a.stringaris@ucl.ac.uk
    affiliations: 
        - id: ucl_1
          name: University College London
          department: Divisions of Psychiatry and Psychology and Language Science
          address: 1-19 Torrington Place
          city: London, UK
          state: 
          postal-code: WC1E 7HB
          country: United Kingdom
    attributes:
        corresponding: true
    note: 
  - name: Charlotte Burman
    email: charlotte.burman@ucl.ac.uk
    affiliations:
        - ref: ucl_1
    note:
  - name: Raphaelle Delpech
    email: raphaelle.delpech.18@ucl.ac.uk
    affiliations:
        - ref: ucl_1
  - name: Rudolf Uher
    email: uher@Dal.Ca
    affiliations:
        - id: dal
          name: Dalhousie University
          department: Department of Psychiatry
          address: 
          city: Nova Scotia
          state: 
          postal-code: 
          country: Canada
  - name: Dayna Bhudia
    email: dayna.bhudia@ucl.ac.uk
    affiliations:
        - ref: ucl_1
  - name: Despoina Miliou
    email: dgmiliou@outlook.com.gr
    affiliations:
        - ref: ucl_1
  - name: Ioannis-Marios Rokas
    email: rokasgiannis93@gmail.com
    affiliations:
        - ref: ucl_1
  - name: Marinos Kyriakopoulos
  - name: Lucy Foulkes
    email: lucy.foulkes@psych.ox.ac.uk
    affiliations:
        - id: OU
          name: University of Oxford
          department: Experimental Psychology
          address: Anna Watts Building, Radcliffe Observatory Quarter, Woodstock Road
          city: Oxford
          postal-code: OX2 6GG
          country: United Kingdom
    note:
  - name: Carmen Moreno
    email: cmoreno@hggm.es
    affiliations:
        - id: HGM
          name: Hospital Gregorio Marañón
          department: Department of Psychiatry
          address: 46 C. del Dr. Esquerdo
          city: Madrid
          state:
          postal-code: 28007
          country: Spain
  - name: Samuele Cortese
    email: samuele.cortese@soton.ac.uk
    affiliations:
        - id: soton
          name: University of Southampton
          department: Centre for Innovation in Mental Health (CIMH), School of Psychology
          address: Highfield Campus, Building 44
          city: Southampton
          state:
          postal-code: SO17 1BJ
          country: United Kingdom
  - name: Glyn Lewis
    email: glyn.lewis@ucl.ac.uk
    affiliations:
        - ref: ucl_1        
  - name: Georgina Krebs
    email: g.krebs@ucl.ac.uk
    affiliations:
        - ref: ucl_1
    note:
keywords: 
  - depression
  - adolescents
  - antidepressants
  - psychotherapy
date: last-modified
format:
  docx:
    reference-doc: "Render_files/custom-reference-doc.docx"
    fig-width: 9
    fig-height: 7
    fig_caption: yes
    tbl_caption: yes
    number-sections: true
execute: 
  echo: true
  auto: true
  warning: FALSE
bibliography: "Render_files/References/emotion_concepts_paper.bib"
csl: "Render_files/References/vancouver.csl"
font:
  family: Calibri
fig-cap-location: top
---

@@

########### Packages needed
library(tidyverse)
library(truncnorm)
library(metafor)
library(meta)
library(kableExtra)
library(flextable)
library(magrittr)
library(officer)
library(RefManageR)
library(ftExtra)
library(dplyr)
library(broom)
library(effsize)
library(here)

@@

@@

**Objectives:** Should a young person receive psychotherapy or medication for their depression, and on what evidence do we base this decision? In this paper, we test whether the basic conditions required to draw valid inferences to answer this question are currently met.

**Methods:** We included 92 RCTs of psychotherapy and medication for child and adolescent depression (mean age 4-18 years). Using meta-analyses, we compared a) participant characteristics and b) trial characteristics in medication and psychotherapy trials. Lastly, we examined whether psychotherapy controls are well-matched to active conditions.

**Results:** Participants in medication RCTs had higher depression severity and were more frequently male compared to psychotherapy RCTs. There was a dramatic difference in the within-subject improvement due to placebo (SMD=-1.9 (95% CI: -2.10 to -1.7)) vs psychotherapy controls (SMD=-0.6 (95% CI: -0.86 to -0.32)). Within psychotherapy RCTs, control conditions were less intensive on average than active conditions.

**Conclusions:** Medication and psychotherapy RCTs differ on fundamental participant and methodological characteristics, thereby violating key conditions for valid comparison between them. Psychotherapy controls often involve little therapist contact and are easy-to-beat comparators. These findings cast doubt on the confidence with which psychotherapy is recommended for youth depression, and highlight the pressing need to improve the evidence base.

@@

**What is already known on this topic**

Psychotherapy is recommended before medication for most cases of depression in children and adolescents, a recommendation that is based on indirect comparisons of outcomes from randomised controlled trials (RCTs) within each treatment modality.

**What this study adds**

We examine the validity of these inferences by scrutinising the comparability of psychotherapy and medication RCTs.

We find significant differences in sample characteristics (namely depression severity and sex composition) and trial design features, such that the within-group effect sizes of medication controls (i.e. pill placebo) are much larger than those for psychotherapy controls, and that medication RCTs feature significantly more trial sites.

We also examine the quality of controls used in psychotherapy RCTs and find that they are poorly matched to active intervention arms in ways such as human contact hours, and hence represent poor and easy to beat comparators.

**How this study might affect research, practice or policy**

Our findings underscore the need for a higher quality evidence base upon which to base treatment guidelines and clinical decision making.

@@

For adolescent depression, there are limited head-to-head trials of medication and psychotherapy, and hence recommendations are derived from indirect comparisons of treatment efficacy. The National Institute of Health and Care Excellence (NICE) guidelines for adolescent depression recommend psychotherapy over medication in most cases.[@nice_depression_2019] This is in keeping with two sources of evidence relating to child and adolescent depression: meta-analyses of medication randomised controlled trials (RCTs) that cast doubt on the efficacy of antidepressants, with the exception of fluoxetine [@cipriani_comparative_2016]; and meta-analyses of psychotherapy RCTs that conclude psychotherapy to be efficacious.[@cuijpers_effects_2021] However, a recent network meta-analysis (NMA),[@zhou_comparative_2020] an established method of comparing treatments using both direct and indirect (i.e. treatment A with treatment C, via studies that directly compare A with B and B with C) evidence, concluded that only fluoxetine alone and fluoxetine administered together with CBT were significantly more effective than medication- (i.e. pill placebo) or psychotherapy-controls. A large head-to-head RCT comparing modalities found that fluoxetine, alone and in combination with CBT, was superior to pill placebo, though CBT alone was not.[@march_fluoxetine_2004] Also, the addition of psychotherapy to standard care did not improve outcomes.[@goodyer_randomised_2008] Given this confusing evidence base, how should we make treatment decisions?

In this paper, we examine whether the existing evidence for adolescent depression treatments can offer valid answers to this question. We provide a conceptual framework and test a series of hypotheses using data from existing trials. Two points are crucial to indirect comparisons of treatment modalities. First, whether the participants in trials are comparable across modalities, or differ in potential effect modifiers. Second, whether key conditions of the trial, such as the effects of control conditions or the number of sites involved, are comparable.

Starting with the first point, comparison between different trials assumes that they sample from populations that are comparable in terms of characteristics that could be effect modifiers. If not, the validity of any comparisons, including those conducted through NMA (which rests on the principal of transitivity, i.e. the requirement that the different sets of randomized trials are similar on average)[@chaimani_chapter_2023] are questionable.

The assumption that medication and psychotherapy trials sample from comparable population may not be valid as patients and parents often have treatment preferences,[@jaycox_adolescent_2006; @langer_parent_2021; @mchugh_patient_2013] meaning that there is likely to be a self-selection bias in who participates in psychotherapy and medication trials. Moreover, treatment preferences correlate with clinically-relevant participant characteristics, including severity and sex. Some of these characteristics, such as severity, may moderate treatment response [@courtney_predictors_2022; @lorenzo-luaces_double_2020] and may confound comparisons.

Regarding the second point, differences in trial design may impact outcomes in a differential way between medication and psychotherapy trials.[@del_giovane_combining_2019] Most obviously, participants in psychotherapy trials are generally unblinded to treatment allocation, with the exception perhaps of trials that compare two equally plausible treatment arms.[@calvo_intervention_2014] This creates differential expectations which may favour the psychotherapy active condition, as participants are content to be receiving the “cutting edge” treatment, whilst those in the control are dissatisfied for having missed out (i.e. “disappointment bias”).[@relton_ethics_2017] By contrast, in new antidepressant trials, patients (and raters) were largely unable to judge treatment allocation,[@lin_assessment_2022] suggesting that expectancy effects are well-matched across conditions. Since expectancy is substantially associated with treatment outcomes,[@constantino_expectations_2011] if expectancy differs between medication and psychotherapy trials, comparisons between them, including in NMA, become questionable.

Another difference in design is the number of trial sites. The number of sites in medication trials is positively related to the magnitude of placebo response.[@bridge_placebo_2009, @dechartres_single-center_2011, @meister_placebo_2020] This phenomenon may be due to lower quality of assessments in multi-site trials, with higher rates of classification errors and therefore higher apparent spontaneous remission or regression to the mean.

@@

Additionally, control conditions in RCTs should generate counterfactual conditions to the intervention: what would have been the outcome had an individual not received the intervention, with all else being equal.[@guo_chapter_2014] Pill placebo, where the appearance of the drug is faithfully emulated, is an effort for all else to be equal. In psychotherapy trials, control conditions may not be so well matched to the intervention (e.g. in number of contact hours).

@@

We examine RCTs of psychotherapy and medication for child and adolescent depression (mean age 4-18 years). We posit there are substantial differences between psychotherapy and medication RCTs, making their comparison problematic and examine the following: First, we conduct meta-analyses to compare sample characteristics of medication and psychotherapy trials including: a) baseline depression severity; b) percentage females; and c) mean age. Second, we examine trial characteristics including the efficacy of the control arms, using random-effects meta-regression, and the number of trial sites. Third, we scrutinise the extent to which psychotherapy controls matched the active intervention in ways such as number and frequency of sessions, and hence whether they represent fair pairings from which to draw valid efficacy inferences.

@@

@@

We included RCTs identified in a recent meta-analysis of psychotherapy versus control,[@cuijpers_effects_2021] an NMA examining the efficacy of antidepressants,[@cipriani_comparative_2016] and an NMA comparing both treatment types [@zhou_comparative_2020] for depression in children and adolescents. For the psychotherapy trials, we utilised open data from the previous meta-analysis.\[@cuijpers_effects_2021 For medication trials, we were unable to access the full dataset used in the NMA and hence extracted data from the included studies ourselves.

For medication trials, we also conducted a systematic search for studies published after the final search date of Cipriani et al.’s [@cipriani_comparative_2016] review up to the final search date of Cuijpers et al’s [@cuijpers_effects_2021] review to ensure we analysed an equivalently up-to-date database of medication trials. Two authors screened 450 titles and abstracts, and 38 full text records. Seven studies met inclusion criteria and one author completed data extraction for these papers.

@@

We conducted random-effects meta-analyses and tested subgroup differences (psychotherapy vs medication trials) in severity of depressive symptoms, sex and age. Meta-analyses were implemented using R’s meta package.

@@

As the measure of effect of each individual study, we used the within-group Standardised Mean Difference (SMD) for the primary depression scale used (selected using the hierarchy in the Supplement).

Where individual studies did not report all data required to calculate the SMD, we imputed missing data according to the methods summarised in this Cochrane Handbook.[@higgins_chapter_2023]

For meta-analysis it is necessary to estimate a standard error of the SMD. This requires a correlation between the pre- and post-measures, a statistic typically not reported. To ensure that our results are not biased by misestimation,[@cuijpers_pre-post_2017] we simulated n=1000 datasets for different values (0.45 to 0.9) of this correlation and used these datasets in subsequent analyses.

@@

We estimated the pooled SMD for each arm by using multilevel models implemented in R’s metafor package.

We present the SMDs of each of the four treatment arms (medication-control, medication-active, psychotherapy-control, psychotherapy-active) under investigation. The SMDs are the means across the 1000 simulated datasets.

@@

We conducted a series of sensitivity analyses where we excluded studies that used waitlist as their control and recruited participants with subclinical levels of depression. Next, we included only trials that used the Children’s Depression Rating Scale, Revised (CDRS-R) or the Hamilton Depression Rating Scale (HAM-D) as outcome instruments. Next, we restricted the analysis to studies with variance below 0.02.

Further, we tested whether the simulated values for the standard error had a substantial influence on the estimation of the differences between the medication and psychotherapy control conditions. We plotted the z-value of the difference between the two coefficients against the number of simulations. We make inference on the stability of the difference, by counting the proportion of times that the z-value is above the critical value of z = 1.645 corresponding to an alpha = 0.05.

@@

Data for included studies are summarised in @tbl-all-trials and available on the project repository (“df_long_for_metan.csv”)[@stringaris_apples_2023].

In total, there were `r length(unique(df_long_for_metan$new_study_id))` RCTs which included `r sum(df_count_studies[df_count_studies$condition == "medication_active", "n"])` active arms and `r sum(df_count_studies[df_count_studies$condition == "medication_control", "n"])` control arms of medication trials; and `r sum(df_count_studies[df_count_studies$condition == "psychotherapy_active", "n"])` active arms and `r sum(df_count_studies[df_count_studies$condition == "psychotherapy_control", "n"])` control arms from psychotherapy RCTs. Note that the number of active and control arms does not match because some studies feature more than one control or active arm.

Placebo pill was the control condition for all medication trials. In psychotherapy trials, the control arm included 14 waitlist, 25 treatment-as-usual (TAU), and 19 other control conditions.

@@

@tbl-baseline-results summarises the results from each of the meta-analyses examining sample characteristics at baseline. The summary statistics are provided for each subgroup and the p-value derives from the test for subgroup differences.

@@

We conducted a series of sensitivity analyses (all figures in the Supplement). Excluding waitlist control studies (see @fig-plot-means-no-wl) and sub-clinical studies (see @fig-plot-means-clin) yielded a pattern of results very similar to the overall analyses. Next, we examined the data including only those studies that used the CDRS (see @fig-plot-means-cdrs) or the HAMD (see @fig-plot-means-hamd). Medication control and psychotherapy control conditions remained significantly different, though the small number of studies resulted in less precise estimates of the SMDs. Restricting the analysis to studies with variance below 0.02 yielded a similar pattern of results, but with increased precision in SMD estimates across conditions (see @fig-plot-means-var2). Finally, we showed that different values for the pre-post measure correlation had minimal effect on the estimated outcomes (see @fig-stab-sims).

@@

We addressed potential regression to the mean by including the baseline score for each depression scale in the linear regression model as per equation 3 in [@barnett_regression_2004] (see @tbl-de-meaned). The difference between the medication control and psychotherapy control arms remained significantly different.

@@

Average number of trial sites was significantly higher in medication trials (*M* = `r df_sites_summary_stats$Mean[df_sites_summary_stats$Modality == "Medication"]`, *SD* =`r df_sites_summary_stats$SD[df_sites_summary_stats$Modality == "Medication"]`) compared to psychotherapy studies (*M* =`r df_sites_summary_stats$Mean[df_sites_summary_stats$Modality == "Psychotherapy"]`, *SD* =`r df_sites_summary_stats$SD[df_sites_summary_stats$Modality == "Psychotherapy"]`)(*t* (`r round (sites_t_test$parameter, 2)`) = `r round(sites_t_test$statistic,2)`, *p* =`r sites_t_test_stats$p_value`). Of those studies with data available, 26 of 28 (93%) medication trials were multisite, compared to 24 of 45 (54%) psychotherapy studies.

@@

As seen in @tbl-intensity-of-int, active conditions featured significantly more sessions when compared to control conditions. Sessions in active conditions were longer and more frequent, resulting in significantly more intervention hours overall. Notably, many control conditions were very poorly described and their intensity could not be quantified, resulting in missing data. We performed a sensitivity analysis where we excluded trials using waitlist controls; with the exception of number of sessions, differences between active and control arms were no longer statistically significant though remained substantial (see @tbl-intensity-of-int-no-wl).

@@

We examined whether psychotherapy and medication can be meaningfully compared on the basis of the existing evidence, by looking at two key conditions for comparability. First, whether the participants of trials in one modality are comparable to those in another modality. Second, whether conditions of the trial, such as the effects of control conditions or the number of sites involved, are comparable.

Starting with the first question, we found that participants in medication trials are comparable on age but are more likely to be male and have more severe depression compared to those in psychotherapy. This indicates that different people enter medication and psychotherapy trials; as these could be effect modifiers, they may violate basic assumptions of comparability.

Severity is particularly important as it may moderate treatment response, with some evidence suggesting that those with higher baseline scores respond more to antidepressants [@stone_response_2022] or that their response to pill placebo is lower.[@bridge_placebo_2009] Other studies argue against severity as a treatment moderator, [@troger_baseline_2024, @weitz_baseline_2015] however these are within people who have chosen to be in the particular trial and modality. Moreover, severity may represent different subtypes in terms of course of depression and real-life outcomes.[@lamers_six-year_2016, @simmonds-buckley_depression_2021]
